---
title: "WISG Module 4: Computational Pipeline for WGS Data"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
link-citations: yes
github-repo: UW-GAC/WISG_2019
description: "Course materials for WISG Module 4"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, message=FALSE)
```

# Introduction

This site contains course materials for WISG Module 4: Computational Pipeline for WGS Data, January 23-24, 2019.
Data used is located in the [github repository](https://github.com/UW-GAC/WISG_2019) from which the site is built, as well as in the [TOPMed analysis pipeline](https://github.com/UW-GAC/analysis_pipeline).

## Schedule

[Lecture slides](https://drive.google.com/drive/folders/1-sp3l6qumvFjohAqMy8WC513PXeHqeiJ?usp=sharing)

**Wednesday, January 23**

- [Introduction](https://drive.google.com/file/d/1TR5QLiKxi5RaXZv-MNGMm-rHS4cMoTHV/view)
- Sequencing and data formats
    - [Sequencing overview](https://drive.google.com/file/d/11lJ3scUzjzdVZCqdP5nXEafZTro6ijIb/view)
    - [Sequencing data formats](https://drive.google.com/file/d/1uz0yS10TWVS1e-8agAqIG6XFtwkU8JVo/view)
    - [Intro to Genomic Data Storage](https://drive.google.com/file/d/1cPAYzGIrd9lPZNhcoiZtFoGEZZO_bqk_/view)
    - [Exercises](https://uw-gac.github.io/WISG_2019/gds-format.html)
- Association tests
    - [Methods and motivation](https://drive.google.com/file/d/1ENC_DeSSr-BHX3aQHlU_3HdW7jk58_Yk/view) (Parts 1-2)
    - [GENESIS for association tests](https://drive.google.com/file/d/1K3BdDMZ75zZPWetOrdt7vMFE7cubFObr/view)
    - [Exercises](https://uw-gac.github.io/WISG_2019/association-tests.html)
    - [Aggregate tests](https://drive.google.com/file/d/1ENC_DeSSr-BHX3aQHlU_3HdW7jk58_Yk/view) (Part 3)
    - [Exercises](https://uw-gac.github.io/WISG_2019/association-tests.html#sliding-window-tests)

**Thursday, January 24**

- Population structure and relatedness
    - [Population structure inference](https://drive.google.com/file/d/1pEbcL6EjGnlzKsrASoP1vOaQnF_dOBFa/view)
    - [Relatedness inference](https://drive.google.com/file/d/1UHNZ72Rk8dDtR2Xp0B8gs7yGCu-rpPxR/view)
    - [R packages for PCA and relatedness](https://drive.google.com/file/d/1VPmkZAZXY-VlIl8GRz28WiBWJfcoSZEZ/view)
    - [Exercises](https://uw-gac.github.io/WISG_2019/computing-a-grm.html)
- [Mixed model association testing](https://drive.google.com/file/d/1dPfsN7lSFJiqzM4bpD5t8zyrYoEW4rtY/view)
    - [Exercises](https://uw-gac.github.io/WISG_2019/mixed-models.html)
- [Variant annotation](https://drive.google.com/file/d/1_gAUYoe2leqJt2ywoNx9caMzMje7YsYp/view)
    - [Exercises](https://uw-gac.github.io/WISG_2019/variant-annotation.html)
- [Analysis pipeline on the cloud](https://drive.google.com/file/d/1HEaa7GRcMYuEmo0yfwp9tTBT6ndQ-sLm/view)


## Software requirements

- [R 3.5.1](https://cran.r-project.org/)
- R packages
    - [SeqArray](http://bioconductor.org/packages/release/bioc/html/SeqArray.html)
    - [SeqVarTools](http://bioconductor.org/packages/release/bioc/html/SeqVarTools.html)
    - [GENESIS](http://bioconductor.org/packages/release/bioc/html/GENESIS.html)
    - [SNPRelate](http://bioconductor.org/packages/release/bioc/html/SNPRelate.html)
    - [TopmedPipeline](https://github.com/UW-GAC/analysis_pipeline/tree/master/TopmedPipeline)
    - dplyr
    - ggplot2
    - GGally
    - survey
    - CompQuadForm
    - TxDb.Hsapiens.UCSC.hg19.knownGene
    
Installation instructions

```{r install, eval=FALSE}
install.packages(c("BiocManager", "remotes"), repos="https://cloud.r-project.org")

BiocManager::install(c("SeqArray", "SeqVarTools", "SNPRelate", "GENESIS", 
                       "survey", "CompQuadForm", "dplyr", "ggplot2", "GGally",
                       "TxDb.Hsapiens.UCSC.hg19.knownGene"))

remotes::install_github("UW-GAC/analysis_pipeline/TopmedPipeline")
```

Download the workshop data and exercises: https://github.com/UW-GAC/WISG_2019/archive/master.zip

## Resources

If you are new to R, you might find the following material helpful:

- [Graphics with ggplot2](http://tutorials.iq.harvard.edu/R/Rgraphics/Rgraphics.html) tutorial
- Data manipulation with [dplyr](http://dplyr.tidyverse.org/)
