# Mixed models

These exercises introduce relatedness to association testing with mixed models.

## Null model

The first step in an association test is to fit the null model. In addition to the `AnnotatedDataFrame` with phenotypes we used previously, we will need the principal components and kinship. We will use the first five PCs to adjust for ancestry.

```{r null_model_mm}
# sample annotation
sampfile <- "data/sample_phenotype_annotation.RData"
annot <- TopmedPipeline::getobj(sampfile)
library(Biobase)
head(pData(annot))

# load the PCs
pcfile <- "data/pcs.RData"
pcs <- TopmedPipeline::getobj(pcfile)
pcs <- pcs[,c("sample.id", "PC1", "PC2", "PC3", "PC4", "PC5")]
head(pcs)

# add PCs to the sample annotation
dat <- left_join(pData(annot), pcs, by="sample.id")
pData(annot) <- dat
save(annot, file="sample_phenotype_pcs.RData")
```

We create a kinship matrix from the output of `pcrelate`. We multiply the kinship values by 2 to get values equivalent to a GRM. This matrix is represented in R as a symmetric matrix object from the Matrix package.

```{r load_kinship}
kinfile <- "data/pcrelate_kinship.RData"
pcrel <- TopmedPipeline::getobj(kinfile)
kinship <- pcrelateToMatrix(pcrel, scaleKin=2, verbose=FALSE)
dim(kinship)
kinship[1:5,1:5]
```

We fit the null model, adding the PCs to the list of covariates, and specifying the kinship as the covariance matrix with the `cov.mat` argument. As before, we use `study` as a grouping variable.

```{r null_model_fit_mm}
library(GENESIS)
nullmod <- fitNullModel(annot, outcome="height", 
                        covars=c("sex", "age", "study", paste0("PC", 1:5)), 
                        cov.mat=kinship, group.var="study", verbose=FALSE)
save(nullmod, file="null_mixed_model.RData")
```

## Single-variant tests

Now we can run a single-variant test, accounting for relatedness between the subjects.

```{r assoc_single_mm}
library(SeqVarTools)
gdsfile <- "data/1KG_phase3_subset_chr1.gds"
gds <- seqOpen(gdsfile)
seqData <- SeqVarData(gds, sampleData=annot)
iterator <- SeqVarBlockIterator(seqData, verbose=FALSE)
assoc <- assocTestSingle(iterator, nullmod)
head(assoc)

qqPlot(assoc$Score.pval)
```



## Exercise

Run a sliding window test using the mixed model and make a QQ plot.


# Mixed models - Solutions

Run a sliding window test using the mixed model and make a QQ plot.

```{r exercise_mm}
seqResetFilter(seqData, verbose=FALSE)
iterator <- SeqVarWindowIterator(seqData, windowSize=5000, windowShift=2000, verbose=FALSE)
assoc <- assocTestAggregate(iterator, nullmod, test="Burden", AF.max=0.1, weight.beta=c(1,1))
head(assoc$results)
head(assoc$variantInfo)

qqPlot(assoc$results$Score.pval)
```

```{r assoc_mm_close}
seqClose(gds)
```
